---
title: "p8105_hw2_cl4469"
author: "Chen Liang"
date: "2023-10-01"
output: github_document
---

```{r load_libraries}
library(tidyverse)
library(readxl)
```

## Peoblem 1

Read and clean the pols-month dataset:
```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols_month_df <- read_csv("data1/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE)|>
  mutate(year = as.numeric(year)) |>
  left_join(x = _, y = month_df) |>
  mutate(president = NA,
         president = ifelse(prez_gop == 0 & prez_dem == 1, "dem",
        ifelse(prez_gop == 1 & prez_dem == 0, "gop", NA))) |>
  select(-prez_dem, -prez_gop,-day) |>
  relocate(year,month, president)
```

Read and clean the snp dataset:
```{r}
snp_df <- read_csv("data1/snp.csv") |>
  janitor::clean_names()|>
  separate(date,
           into=c("day","month","year"),
           sep="/",convert=T) |>
  mutate(year=ifelse(year>50,1900+year, year+2000),
         month=month.name[month]) |>
  select(-day) |>
  relocate(year, month)
```

Read and clean the unemployment dataset:
```{r}
unemployment = 
  read_csv("data1/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Merging snp_df into pols_month_df, and merging unemp_df into the result
```{r}
pols_snp_df=left_join(pols_month_df,snp_df,by=c("year","month"))
data_538=(left_join(pols_snp_df,unemployment,by=c("year","month")))
str(data_538)
```

## Problem 2

```{r}
library(tidyverse)
library(readxl)
```

#### Reading and cleaning the Mr. Trash Wheel sheet:
```{r}
trash_path = "data2/202309 Trash Wheel Collection Data.xlsx"
mr_trash = 
  read_excel(
  path = trash_path,
  sheet = "Mr. Trash Wheel",
  range = cell_cols("A:N"))|>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(homes_powered=weight_tons*(500/30),
         name="Mr.",
         year = as.numeric(year)) |>
  relocate(name)
```
Description of mr_trash:
The data contains `r nrow(mr_trash)` observations and `r ncol(mr_trash)` variables.
There are several key variables including dumpster number, date of collection,
amount of total litter, types of litter, etc. And a new `homes_powered` variable is
based on the calculation described in the Homes powered note. Then Converting the
character year variable into a numeric one and adding an additional variable `name`
to help with the final merge of the datasets

#### Reading, cleaning and manipulating Professor Trash Wheel sheets:
```{r}
trash_path = "data2/202309 Trash Wheel Collection Data.xlsx"
Prof_df= 
  read_excel(
  path = trash_path,
  sheet = "Professor Trash Wheel",
  range = cell_cols("A:M")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(homes_powered=weight_tons*(500/30),
         name="Prof") |>
  relocate(name)
```
Description of mr_trash:
The data contains `r nrow(Prof_df)` observations and `r ncol(Prof_df)` variables.
There are several key variables including dumpster number, date of collection,
amount of total litter, types of litter, etc. And a new `homes_powered` variable is
based on the calculation described in the Homes powered note. Then adding an
additional variable `name` to help with the final merge of the datasets

#### Reading, cleaning and manipulating Gwynnda sheets:
```{r}
trash_path = "data2/202309 Trash Wheel Collection Data.xlsx"
Gwynnda_df= 
  read_excel(
  path = trash_path,
  sheet = "Gwynnda Trash Wheel",
  range = cell_cols("A:K")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(homes_powered=weight_tons*(500/30),
  name="Gwy") |>
  relocate(name)
```
Description of mr_trash:
The data contains `r nrow(Prof_df)` observations and `r ncol(Prof_df)` variables.
There are several key variables including dumpster number, date of collection,
amount of total litter, types of litter, etc. And a new `homes_powered` variable is
based on the calculation described in the Homes powered note. Then adding an
additional variable `name` to help with the final merge of the datasets

#### Combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset
```{r}
Total_df =
  bind_rows(Prof_df,Gwynnda_df,mr_trash) |>
  janitor::clean_names()
```

#### calculate the total weight of trash collected by Professor Trash Wheel, and the total number of cigarette butts collected by Gwynnda in July of 2021
```{r}
sum(Total_df %>% filter(name == "Prof") %>% select(weight_tons))

sum(Total_df %>% filter(name=="Gwy",year==2021, month=="July") %>% select(cigarette_butts))
```

For the resulting dataset `Total_df`, it has `r nrow(Total_df)` observations and `r ncol(Total_df)`
variables and tells us about how much trash in each of their dumpsters were collected by Mr. Trash
Wheel, professor Trash Wheel, and Gwynnda Trash Wheel between the year of
`r range(Total_df$year[1])` to `r range(Total_df$year[2])`. In detail, the dataset have many key 
variables, such as`weight_tons`,`volume_cubic_yards`,`homes_powered`. The variable `weight_tons` 
measures how much trash were collected, such as `plastic_bottles`,`polystyrene`,`cigarette_butts`.
And the total weight of trash collected by Professor Trash Wheel is 
`r sum(Total_df %>% filter(name == "Prof") %>% select(weight_tons)) `.
At last, In July 2021, the total number of cigarette butts collected by Gwynnda is 
`r sum(Total_df %>% filter(name=="Gwy",year==2021, month=="July") %>% select(cigarette_butts))`.

## Problem 3

#### Read and clean the baseline_df dataset:
```{r}
baseline_df <- read_csv("data3/MCI_baseline.csv", skip=1)|>
  janitor::clean_names() |>
  mutate(sex = ifelse(sex ==1, "Male","Female"),
         apoe4= ifelse(apoe4==1,"Carrier","Non-carrier"),
         age_at_onset=as.numeric(age_at_onset))|>
  filter(current_age < age_at_onset | is.na(age_at_onset))
```
The original dataset `baseline_df`, it has `r nrow(baseline_df)` observations and
`r ncol(baseline_df)`variables, which describes the age at the study baseline, sex, years of
education, apoe4 carry status and age at the onset of MCI. First, I clean and tidy data and use
mutate function to change the `sex` and `apoe4` to non-numeric variable types. Second, I
converted the age_at_onset variable from character to numeric so that the dataset can be
filtered in the next step. At last, If the patients `current_age` is larger than`age_at_onset`,
these participants are not meet the criteria because they have MCI at baseline, and these
observations were removed, so I filtered to exclude patients who do not have MCI at baseline.
After this process, there are `r nrow(baseline_df)` observations left.

```{r}
#How many participants were recruited
  nrow(baseline_df)
#of these how many develop MCI
  baseline_clean = baseline_df %>%filter(!is.na(age_at_onset))
  nrow(baseline_clean)
#What is the average baseline age? 
  mean(baseline_df$current_age)
#What proportion of women in the study are APOE4 carriers?
##The number of females:  
  nrow (filter(baseline_df, sex == "Female"))
##The number of females who have apeo4 carrier:
  nrow(filter(baseline_df, sex == "Female" & apoe4 == "Carrier"))
#The ratio
  nrow(filter(baseline_df, sex == "Female" & apoe4 == "Carrier"))/nrow (filter(baseline_df, sex == "Female"))
```
For the original dataset `baseline_df`, there are a total of `r nrow(baseline_df)` participants recruited. Among this participants, `r nrow(baseline_clean)` participants develop MCI.The average age of the patients was `r mean(baseline_df$current_age)` years old. The dataset contained `r nrow (filter(baseline_df, sex == "Female"))` females. Among these women, `nrow(filter(baseline_df, sex == "Female" & apoe4 == "Carrier"))` were carriers. Thus, the proportion of women in the study are APOE4 carriers is `r nrow(filter(baseline_df, sex == "Female" & apoe4 == "Carrier"))/nrow (filter(baseline_df, sex == "Female"))`

```{r}
amyloid_df <- read_csv("data3/mci_amyloid.csv",skip=1) |>
  janitor::clean_names()|>
  rename(id=study_id,
         time_0=baseline) |>
  pivot_longer(
    time_0:time_8, 
    names_to = "years_since_baseline",
    values_to = "baseline")
```
The `amyloid_df` dataset contains the data of `r nrow(amyloid_df)` patients and `r ncol(amyloid_df)`columns. First, I renamed the columns to be more descriptive of the variable by using `r rename(id = study_id, time_0=baseline)`. Second, because the `amyloid_df` dataset has longitudinally observed biomarker values. For this dataset, I have multiple columns `time_0`,`time_2`,`time_3`representing different variables, and I want to stack those columns into a single variable column`years_since_baseline`. Thus,I need to use `pivot_longer`function to tidy it.

```{r}
baseline=baseline_df |>
anti_join(amyloid_df, by= "id")
nrow(baseline)
  
amyloid=amyloid_df |>
anti_join(baseline_df, by= "id")
nrow(amyloid)/5

baseline=baseline_df |> anti_join(amyloid_df, by= "id") nrow(baseline)
Combine_df = inner_join(baseline_df, amyloid_df, by = "id")
write.csv(Combine_df, file = "data3/Combine_df.csv")
```
Then, I combined the baseline_df and amyloid_df datasets so that only participants who appear in both datasets are retained. The `Combine_df` dataset contains `r nrow(Combine_df)` rows and `r ncol(Combine_df)` columns.The `Combine_df` has `r nrow (filter(Combine_df, sex == "Female"))` females and `r nrow (filter(Combine_df, sex == "Male"))` males. Among them, There are `r nrow(filter(Combine_df, apoe4 == "Carrier"))` carriers.The average age of the patients is `r mean(Combine_df$current_age)` years old.The average years of education of the patients is `r mean(Combine_df$education)` years. The average age of onset of MCI for the patients is `r mean(Combine_df$age_at_onset)` years.

Moreover, there are `r `nrow(baseline)` patients who are in the `baseline_df` dataset and not in the `amyloid_df` dataset. There are `r nrow(amyloid)/5` patients who are in the `amyloid_df` dataset and not in the `baseline_df` dataset.

In the end, I exported the dataset as a csv file.
